---
title: "Causal simulations"
format:
  html:
    df-print: kable
---

## Setup environment

```{r}
#| output: false
library(coursekata)

set.rseed(666)

theme_set(theme_minimal())

n <- 1000
```

## Helper functions

```{r}
rbern <- function(n, prob = 0.5) rbinom(n, size = 1, prob)
```

## Ice cream sales and shark attacks

### Generate data with confounding by season (summer)

```{r}
summer <- rbern(n)

ice_cream_sales <- rbern(n, prob = 0.1 + 0.2 * summer)
shark_attacks   <- rbern(n, prob = 0.01 + 0.02 * summer)

df <- tibble(shark_attacks, ice_cream_sales, summer)
```

```{r}
lm(shark_attacks ~ ice_cream_sales, data = df)
```

```{r}
lm(shark_attacks ~ ice_cream_sales + summer, data = df)
```

### Joint probability table for ice cream sales and shark attacks

```{r}
joint_counts <- tally( ~ ice_cream_sales + shark_attacks, format = "data.frame")

joint_counts
```

```{r}
joint_p <- joint_counts %>% mutate(p = Freq / sum(Freq)) %>% select(-Freq)

joint_p
```

```{r}
gf_tile(shark_attacks ~ ice_cream_sales, fill = ~ p, data = joint_p) %>%
  gf_refine(scale_fill_gradient(low = "white", high = "blue", limits = c(0, 1))) %>%
  gf_labs(title = "Joint probability distribution of Ice Cream Sales and Shark Attacks", x = "Ice Cream Sales", y = "Shark Attacks")
```

### Calculate marginal and conditional probabilities

**Marginal probability of shark attacks**

```{r}
marginal_shark_attacks <- joint_p %>%
  group_by(shark_attacks) %>%
  summarize(`P(shark_attacks)` = sum(p)) %>%
  ungroup()

marginal_shark_attacks
```

**Conditional probabilities of shark attacks given ice cream sales**

```{r}
cond_shark_attacks_given_ice_cream_sales <- joint_p %>%
  group_by(ice_cream_sales) %>%
  mutate(`P(shark_attacks | ice_cream_sales)` = p / sum(p)) %>%
  ungroup() %>% 
  select(shark_attacks, everything()) %>% 
  arrange(ice_cream_sales)

cond_shark_attacks_given_ice_cream_sales
```

**Marginal probability of ice cream sales**

```{r}
marginal_ice_cream_sales <- joint_p %>%
  group_by(ice_cream_sales) %>%
  summarize(`P(ice_cream_sales)` = sum(p)) %>%
  ungroup()

marginal_ice_cream_sales
```

**Conditional probabilities of ice cream sales given shark attacks**

```{r}
cond_ice_cream_sales_given_shark_attacks <- joint_p %>%
  group_by(shark_attacks) %>%
  mutate(`P(ice_cream_sales | shark_attacks)` = p / sum(p)) %>%
  ungroup() %>% 
  select(ice_cream_sales, everything()) %>% 
  arrange(shark_attacks)

cond_ice_cream_sales_given_shark_attacks
```

## Elemental confounds

```{r}
library(tidyverse)
library(dagitty)
library(ggdag)

theme_set(theme_minimal())
```

```{r}
coords <- list(x = c(X = 0, Y = 2, Z = 1), y = c(X = 0, Y = 0, Z = 1))

fork <- dagify(
  X ~ Z,
  Y ~ Z,
  exposure = "X",
  outcome = "Y",
  coords = coords
)

chain <- dagify(
  Z ~ X,
  Y ~ Z,
  exposure = "X",
  outcome = "Y",
  coords = coords
)

collider <- dagify(
  Z ~ X + Y,
  exposure = "X",
  outcome = "Y",
  coords = coords
)
```

### Fork

```{r}
ggdag(fork, text = TRUE) + theme_dag()
```

### Chain

```{r}
ggdag(chain, text = TRUE) + theme_dag()
```

### Collider

```{r}
ggdag(collider, text = TRUE) + theme_dag()
```
