---
title: "Bayesian inference workflow" 
subtitle: "Statistical inference"
date: 2025-02-24
author: "Edoardo \"Dado\" Marcora"
format:
  revealjs:
    smaller: true
    incremental: true
    theme: [default, styles.scss]
    execute:
      echo: true
      eval: true
      warning: false
      fragment: true
    output-location: fragment
df-print: kable
---

```{r}
#| include: false
library(tidyverse)
library(ggformula)
theme_set(theme_minimal())

#| output-location: default
# def conditional operator
`%=>%` <- function(A, B) {
  !A | B
}

# def biconditional operator 
`%<=>%` <- function(A, B) {
  (A %=>% B) & (B %=>% A)
}
```

```{r}
#| include: false
library(rlang)
```

```{r}
#| include: false
P <- function(..., H) {
  # capture premises as a list of quosures
  premises <- enquos(...)

  # capture conclusion (H) as a quosure
  H <- enquo(H)

  if (length(premises) == 0) {
    # if no premises are provided, set evidence (E) to TRUE (tautology)
    E <- expr(TRUE)
  } else {
    # else set evidence (E) to conjunction of all premises
    E <- reduce(premises, ~ expr((!! .x) & (!! .y)))
  }
    
  # extract atomic variables from E and H
  vars <- unique(c(all.vars(E), all.vars(H)))
  
  # build truth table and calculate proportion of rows where E is true, in which H is also true
  expand_grid(!!!set_names(rep(list(c(TRUE, FALSE)), length(vars)), vars)) %>%
    mutate(E = !!E, H = !!H) %>%
    filter(E) %>%
    pull(H) %>%
    mean()
}
```

```{r}
#| include: false
library(flextable)

P.flex <- function(..., H, table = TRUE, flex = TRUE) {
  # capture premises as a list of quosures
  premises <- enquos(...)
  
  # capture conclusion (H) as a quosure
  H <- enquo(H)
  
  # extract atomic variables from premises and H
  vars <- unique(c(unlist(lapply(premises, all.vars)), all.vars(H)))
  
  # create data frame with all possible combinations of atomic variables/worlds (universe)
  tt <- expand_grid(!!!set_names(rep(list(c(TRUE, FALSE)), length(vars)), vars))

  # add columns for premises
  for (i in seq_along(premises)) {
    tt[[paste0("P", i)]] <- eval_tidy(premises[[i]], data = tt)
  }
  
  # add columns for conjunction of all premises (E) and H
  tt <- tt %>%
    mutate(
      E = if_all(starts_with("P"), identity),
      H = eval_tidy(H, data = tt)
    )
  
  # calculate P(H | E) by checking whether in every possible world/row where E is true, H is also true
  P <- tt %>%
    filter(E) %>%
    pull(H) %>%
    mean()
  
  # return P(H | E) if truth table output is not requested
  if (!table) {
    return(P)
  }
  
  # return truth table if formatted truth table output is not requested
  if (!flex) {
    return(tt)
  }
  
  # define colors
  highlight_color <- "beige"  # Standard HTML color name
  true_color <- "darkgreen"
  false_color <- "darkred"
  
  # convert the truth table into a flextable
  tt.formatted <- flextable(tt) %>%
    bold(part = "header") %>%  # Bold the header
    bold(part = "footer") %>%  # Bold the footer
    bg(i = which(tt$E == TRUE), bg = highlight_color, part = "body")  # Highlight E = TRUE rows
  
  # apply text colors
  for (col in colnames(tt)) {
    tt.formatted <- tt.formatted %>%
      color(i = which(tt[[col]] == TRUE), j = col, color = true_color) %>%
      color(i = which(tt[[col]] == FALSE), j = col, color = false_color)
  }
  
  # add footer with the calculated P(C | K) value
  tt.formatted <- tt.formatted %>%
    add_footer_row(
      values = paste("P(H | E) =", round(P, 3)),
      colwidths = ncol(tt)
    ) %>%
    align(align = "right", part = "footer")

  # return formatted truth table
  return(tt.formatted)
}
```

##  {background-image="images/last_week_tonight.jpg" background-size="contain"}

## Recap slides

##  {background-image="images/bayesian_inference_is_just_counting.jpg" background-size="contain"}

## Bayesian inference workflow

XXX

## Problem: Multiple draws with replacement from bag of balls

```{r}
bag <- c("B", "W", "W", "W")

ss <- expand_grid(
  D1 = bag
)

ss
```

```{r}
ss %>%
  mutate(
    E = TRUE,
    H = D1 == "B"
  ) %>%
  filter(E) %>% 
  pull(H) %>% 
  mean()
```

```{r}
bag <- c("B", "W", "W", "W")

ss <- expand_grid(
  D1 = bag,
  D2 = bag,
  D3 = bag)

ss
```

```{r}
ss %>%
  mutate(
    E = TRUE,
    H = (D1 == "B" & D2 == "W" & D3 == "B")
  ) %>%
  filter(E) %>% 
  pull(H) %>% 
  mean()
```

##  {background-image="images/thats_all_folks.jpg" background-size="50%"}

##  {background-image="images/briggs2.png" background-size="contain"}
