---
title: 'Marginal effects using R'
date: 2025-03-24
author: Edoardo \'Dado\' Marcora
execute:
  warning: false
format:
  html:
    toc: true
    df-print: kable
bibliography: references.bib
---

## Introduction

This notebook shows how to use various packages to 1) import data from an Excel spreadsheet ([`readxl`](https://readxl.tidyverse.org/)), 2) evaluate and compare models ([`performance`](https://easystats.github.io/performance/) and `modelsummary`), and 3) estimate marginal effects ([`modelbased`](https://easystats.github.io/modelbased/) using the model with interaction (`m5`) of the mouse stress/sociability/neuronal activation experiment illustrated during the lecture of week 11 (2025-03-24, **Chapter 9: Models with interactions**) by [Evan Schaffer](https://www.schafferlab.com/evan-schaffer).

## Setup environment

```{r}
#| output: false
library(tidyverse)
library(ggformula)
library(readxl)
library(performance)
library(modelsummary)
library(modelbased)

theme_set(theme_bw())

set.seed(666)
```

## Background information

The data are from a study of the effects of footshock stress on social behavior and neuronal activation in the amygdala (and medial prefrontal cortex) of male and female mice.

Data available at <https://doi.org/10.7910/DVN/1RAQON>

Study available at <https://doi.org/10.1371/journal.pone.0281388>

## Estimands

The aim of the original study was to explore the effect of stress on the sociability of male and female mice in a standard social approach test. Neuronal activation during social approach was assessed using immunohistochemistry against the immediate early gene product cFos.

Here we are going to estimate the causal effect of stress (`StressCondition`, treatment variable) on neuronal activation (`CeLcfos`, outcome variable) conditional on the sociability of male and female mice (`Sociability`, control variable).

## Read and prep data

This study is a good example of [how NOT to format supplementary data in an Excel spreadsheet]{.underline}! It shouldn't take more than one call to the `read_xlsx()` function to import data from a properly formatted Excel spreadsheet.

```{r}
id1 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!C4:D11', col_names = c('ID', 'Treatment'))
id2 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!C16:D23', col_names = c('ID', 'Treatment'))
id3 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!C26:D33', col_names = c('ID', 'Treatment'))
id4 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!C37:D44', col_names = c('ID', 'Treatment'))

id <- rbind(id1, id2, id3, id4)
```

```{r}
cel1 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!H4:H11', col_names = 'CeLcfos')
cel2 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!H16:H23', col_names = 'CeLcfos')
cel3 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!H26:H33', col_names = 'CeLcfos')
cel4 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!H37:H44', col_names = 'CeLcfos')

cel <- rbind(cel1, cel2, cel3, cel4)
```

```{r}
soc1 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!N4:N11', col_names = 'Sociability')
soc2 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!N16:N23', col_names = 'Sociability')
soc3 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!N26:N33', col_names = 'Sociability')
soc4 <- read_xlsx('Figure 3.xlsx', range = 'New_Beh+cfos!N37:N44', col_names = 'Sociability')

soc <- rbind(soc1, soc2, soc3, soc4)
```

```{r}
d <- cbind(id,soc,cel) %>%
  mutate(Sex = factor(substr(ID, 1, 1))) %>% 
  rowwise() %>% 
  mutate(
    StressCondition = case_when(
      grepl('FC-', Treatment) ~ 'Control',
      grepl('FC+', Treatment) ~ 'Shock',
      grepl('Control', Treatment) ~ 'Control',
      grepl('Stress', Treatment) ~ 'Shock'),
    Treatment = case_when(
      grepl('Veh', Treatment) ~ 'Vehicle',
      grepl('Alp', Treatment) ~ 'Alprazolam')) %>%
  ungroup() %>% 
  mutate(logCeLcfos = log(CeLcfos),
         Sex = factor(Sex),
         Treatment = factor(Treatment),
         StressCondition = factor(StressCondition)) %>% 
  select(ID, CeLcfos, logCeLcfos, StressCondition, Sociability, everything())

head(d)
```


## Causal model

The causal estimand is the effect of stress (`StressCondition`, treatment variable) on neuronal activation (`CeLcfos`, outcome variable). We are going to ignore the `Treatment` and `Sex` variables in this example used to illustrate a simple linear model with two interacting predictors (`StressCondition` and `Sociability`). However, in real-life, you should include them in the causal model, unless of course your assumption is that they are in no way causally related to the exposure and/or the outcome (which is the causal assumption for omitted variables).

Based on the experimental design and my domain knowledge, my DAG is:

```{r}
library(dagitty)

dag <- dagitty('dag {
  CeLcfos [outcome]
  StressCondition [exposure]
  Sociability -> CeLcfos
  StressCondition -> CeLcfos
  StressCondition -> Sociability
}')

coordinates(dag) <- list(
  x = c(StressCondition = 0, Sociability = 0.5, CeLcfos = 1),
  y = c(StressCondition = 0, Sociability = 1, CeLcfos = 0)
)

plot(dag)
```

According to this causal graph, in order to estimate the **direct** causal effect of `StressCondition` on `CeLcfos`, I must include `Sociability` as a covariate in the regression model, in order to block the **indirect** causal effect of `StressCondition` on `CeLcfos` mediated by `Sociability`.

## Exploratory data analysis

```{r}
gf_point(CeLcfos ~ Sociability, color = ~ StressCondition, data = d) %>% gf_smooth(method = 'gam', formula = y ~ s(x, bs = 'cs'), se = TRUE)
```

### CeLcfos \~ NULL

```{r}
gf_point(CeLcfos ~ 1, data = d) %>% coursekata::gf_model(CeLcfos ~ 1, data = d)
```

It is often but erroneously believed that linear regression assumes that the outcome variable is normally distributed. This misconception arises from not knowing/remembering the regression likelihood equation(s) of the general linear model, e.g.:

$$
y_i \stackrel{\text{i.i.d.}} \sim \text{Normal}(\mu = \beta_0 + \beta_1 x_i, \sigma)
$$

or (equivalently):

$$
y_i = \beta_0 + \beta_1 x_i + \epsilon_i
$$

$$
\epsilon_i \stackrel{\text{i.i.d.}} \sim \text{Normal}(0, \sigma)
$$

All the assumptions of the general linear model are unambiguously stated by these two equivalent notations!

Specifically, to return to the misconception mentioned above, the linear regression model assumes that the outcome variable $y$ is normally distributed **conditional on the covariates** $x$ or, equivalently, that the **residuals** **are normally distributed**.

For example:

```{r}
nsamples = 1000

y_xA = rnorm(nsamples, mean = 2, sd = 4)
y_xB = rnorm(nsamples, mean = 20, sd = 4)

y_x <- tibble(y = c(y_xA, y_xB), x = factor(c(rep('A', nsamples), rep('B', nsamples))))
```

```{r}
gf_dens(~ y, data = y_x)
```

```{r}
gf_dens(~ y | x, data = y_x)
```

```{r}
gf_dens(~ y, color = ~ x, data = y_x)
```

```{r}
m <- lm(y ~ x, data = y_x)

summary(m)
```

```{r}
check_normality(m)
```

```{r}
gf_jitter(y ~ x, color = ~ x, alpha = 0.1, data = y_x) %>%
  coursekata::gf_model(m) %>%
  ggExtra::ggMarginal(y, type='density')
```

The 'problem' with the `CeLcfos` outcome variable is not that it is not normally distributed, but that it is a count variable (or more precisely in this case a density, the number of 'active neurons'/cfos-expressing cells per mm^2^ of the central lateral amygdala) and thus it cannot be negative.

A normally distributed variable is defined over the entire real line $(-\infty, \infty)$ and, therefore, it is not appropriate for count data, especially if they are concentrated close to the zero boundary.

```{r}
mosaic::favstats(~ CeLcfos, data = d)
```

```{r}
gf_histogram(~ CeLcfos, data = d)
```

```{r}
gf_qq(~ CeLcfos, data = d) %>% gf_qqline()
```

A log transformation is often applied to count data in order to make them more amenable to commonly used null-hypothesis significance tests and general linear regression modeling, all of which assume that the outcome variable is conditionally normally distributed.

```{r}
mosaic::favstats(~ logCeLcfos, data = d)
```

```{r}
gf_histogram(~ logCeLcfos, data = d)
```

```{r}
gf_qq(~ logCeLcfos, data = d) %>% gf_qqline()
```

However, generalized linear models that, e.g., use the Poisson or negative binomial distribution are better suited for dealing with count data.

The log transformation of counts is also problematic when the data includes observations of zero counts and the whole dataset need to be fudged by adding a value, usually 1, before transformation.

### log(CeLcfos) \~ NULL

```{r}
gf_point(logCeLcfos ~ 1, data = d) %>% coursekata::gf_model(logCeLcfos ~ 1, data = d)
```

### log(CeLcfos) \~ StressCondition

```{r}
gf_point(logCeLcfos ~ StressCondition, color = ~ StressCondition, data = d) %>% coursekata::gf_model(logCeLcfos ~ StressCondition, data = d)
```

### log(CeLcfos) \~ Sociability

```{r}
gf_point(logCeLcfos ~ Sociability, data = d) %>% coursekata::gf_model(logCeLcfos ~ Sociability, data = d)
```

### log(CeLcfos) \~ StressCondition + Sociability

```{r}
gf_point(logCeLcfos ~ Sociability, color = ~ StressCondition, data = d) %>%
  coursekata::gf_model(logCeLcfos ~ Sociability + StressCondition, data = d) %>% gf_refine(theme(legend.position = 'none'))
```

### log(CeLcfos) \~ StressCondition \* Sociability

```{r}
gf_point(logCeLcfos ~ Sociability, color = ~ StressCondition, data = d) %>%
  coursekata::gf_model(logCeLcfos ~ Sociability * StressCondition, data = d) %>% gf_refine(theme(legend.position = 'none'))
```

## Build and fit models

### log(CeLcfos) \~ NULL

```{r}
m1 <- lm(log(CeLcfos) ~ NULL, data = d)

summary(m1)
```

```{r}
#| fig-height: 10
check_model(m1)
```

```{r}
check_autocorrelation(m1)
```

### log(CeLcfos) \~ StressCondition

```{r}
m2 <- lm(log(CeLcfos) ~ StressCondition, data = d)

summary(m2)
```

```{r}
#| fig-height: 10
check_model(m2)
```

```{r}
check_autocorrelation(m2)
```

### log(CeLcfos) \~ Sociability

```{r}
m3 <- lm(log(CeLcfos) ~ Sociability, data = d)

summary(m3)
```

```{r}
#| fig-height: 10
check_model(m3)
```

```{r}
check_autocorrelation(m3)
```

### log(CeLcfos) \~ StressCondition + Sociability

```{r}
m4 <- lm(log(CeLcfos) ~ StressCondition + Sociability, data = d)

summary(m4)
```

```{r}
#| fig-height: 10
check_model(m4)
```

```{r}
check_autocorrelation(m4)
```

### log(CeLcfos) \~ StressCondition \* Sociability

```{r}
m5 <- lm(log(CeLcfos) ~ StressCondition * Sociability, data = d)

summary(m5)
```

```{r}
#| fig-height: 10
check_model(m5)
```

```{r}
check_autocorrelation(m5)
```

## Model evaluation and comparison

Here we use the `modelsummary()` and the `compare_performance()` functions from the [`modelsummary`](https://modelsummary.com/) and [`performance`](https://easystats.github.io/performance/) packages, respectively, to evaluate and compare models.

```{r}
modelsummary(list(m1, m2, m3, m4, m5), stars = TRUE, statistic = 'conf.int')
```

```{r}
plot(compare_performance(m1, m2, m3, m4, m5, rank = TRUE, verbose = FALSE))
```

## Model predictions

Statistical models often contain a lot more insights than what you get from simply looking at the model parameters. In many cases, like models with multiple interactions, non-linear effects, non-standard families, complex random effect structures, the parameters can be hard to interpret.

Except for very simple models, parameters cannot be easily interpreted as quantities of interest. Instead, we can use packages like `marginaleffects`, `emmeans`, or `modelbased` (which is a wrapper around `marginaleffects` and `emmeans` to make them a bit easier to use) to compute and plot predictions, estimate quantities of interest, and perform hypothesis tests with these quantities.

Please note that model predictions are transformed back by exponentiation to the original scale of the outcome variable `CeLcfos` (count of active neurons/mm\^2) for better interpretability.

```{r}
estimate_means(m5, by = c('Sociability', 'StressCondition'), transform = exp) %>% plot()
```

```{r}
estimate_means(m5, by = c('Sociability', 'StressCondition'), transform = exp)
```


## Marginal effects and hypothesis tests

Please note that exponentiating the difference of two logged values is equivalent to dividing the original values, hence the difference here is the ratio/fold-change of shock over control.

```{r}
estimate_contrasts(m5, contrast = 'StressCondition', transform = exp)
```

```{r}
estimate_contrasts(m5, contrast = 'StressCondition', by = 'Sociability', transform = exp)
```

## References

::: {#refs}
:::

## Print environment

```{r}
sessioninfo::session_info()
```
