---
title: "rethinking_lm"
format:
  html:
    toc: true
    df-print: paged
---

## Setup environment

```{r}
#| output: false
library(coursekata)
library(tidyverse)
library(easystats)
library(rethinking)
library(rstanarm)
library(bayesplot)

theme_set(theme_bw())

set.rseed(666)
```

## Load data

```{r}
data(Howell1) # from rethinking package
```

Demographic data from [Kalahari !Kung San people](https://en.wikipedia.org/wiki/%C7%83Kung_people) collected by [Nancy Howell](https://medium.com/extralife/the-long-ceiling-fddda72f4c0b)

-   `height`: Height in cm
-   `weight`: Weight in kg
-   `age`: Age in years
-   `male`: Gender indicator

## Explore data

```{r}
Howell1
```

```{r}
skimr::skim(Howell1)
```

```{r}
gf_point(weight ~ height, data = Howell1, alpha = 0.5, color = 4) %>%
  gf_smooth(linewidth = 1, color = 2, method = "loess") %>% 
  gf_labs(x = "Height (cm)", y = "Weight (kg)")
```

```{r}
d.pop <- Howell1 %>% filter(age >= 18) %>% select(-c(male, age))

d.pop
```

```{r}
skimr::skim(d.pop)
```

```{r}
gf_point(weight ~ height, data = d.pop, alpha = 0.5, color = 4) %>%
  gf_smooth(linewidth = 1, color = 2, method = "loess") %>% 
  gf_labs(x = "Height (cm)", y = "Weight (kg)")
```

## Workflow

1.  State a clear question/goal
2.  Specify causal model (use a DAG to sketch your causal assumptions)
3.  Use causal model to build generative model
4.  Specify estimand (i.e., the unknown statistical quantity of interest)
5.  Use generative model to build an estimator
6.  Test estimator using data simulated with generative model
7.  Estimate (fit/train model using observed data)
8.  Check model assumptions
9.  Test model using training data (posterior predictive check)
10. Test model using new data (replication)
11. Profit but keep testing

## State a clear question/goal

-   What is the relationship between height and weight in adults?

-   Describe the association between height and weight in adults

## Specify causal model (DAG)

```{mermaid}
graph LR;
    H --> W;
```

$$\text{W} = f(\text{H})$$

## Use causal model to build generative model

The likelihood of the general linear model:

$$
y_i \stackrel{\text{i.i.d.}} \sim \text{Normal}(\mu = (\beta_0 + \beta_1 \cdot x_i), \sigma)
$$

or (equivalently):

$$
y_i = \beta_0 + \beta_1 \cdot x_i + \epsilon_i
$$ $$
\epsilon_i \stackrel{\text{i.i.d.}} \sim \text{Normal}(0, \sigma)
$$

```{r}
sim_weight <- function(H, beta1, sigma) {
  U <- rnorm(length(H), mean = 0, sd = sigma)
  W <- beta1 * H + U
  return(W)
}
```

```{r}
H <- rnorm(n = 350, mean = 155, sd = 8)
W <- sim_weight(H, beta1 = 0.4, sigma = 5)
```

```{r}
d.sim <- tibble(height = H, weight = W)
```

```{r}
gf_point(weight ~ height, data = d.sim, alpha = 0.5, color = 3) %>%
  gf_smooth(linewidth = 1, color = 2, method = "loess") %>%
  gf_labs(x = "Height (cm)", y = "Weight (kg)")
```

## Specify estimand

The unknown statistical quantity of interest

## Use generative model to build an estimator

Build Bayesian estimator using grid approximation

```{r}
sample.size <- 20 # number of samples in the dataset
n <- 1000         # number of samples used in Montecarlo simulations
```

```{r}
# parameter variables
beta0 <- seq(-80, 80, length.out = 100)
beta1 <- seq(-3, 3, length.out = 100)
sigma <- seq(0.1, 20, length.out = 100)

# build sample space of parameter variables
ss <- expand_grid(beta0, beta1, sigma)

# simulate data
H <- rnorm(sample.size, mean = 155, sd = 8)
W <- sim_weight(H, beta1 = 0.4, sigma = 5)

# estimate using grid approximation
post <- ss %>%
  rowwise() %>%
  mutate(
    like = exp(sum(dnorm(W, mean = beta0 + beta1 * H, sigma, log = TRUE))),
    post.unnorm = like
  ) %>%
  ungroup() %>%
  mutate(
    post = post.unnorm / sum(post.unnorm)
    
  )

post
```

## Test estimator

$$
\beta_0 = 0 \quad \beta_1 = 0.4 \quad \sigma = 5
$$

```{r}
draws <- post %>%
  slice_sample(n = n, weight_by = post, replace = TRUE) %>%
  select(beta0, beta1, sigma) %>% 
  posterior::as_draws_df()
```

```{r}
mcmc_hist(draws, pars = "beta1", binwidth = 6/100)
```

```{r}
summary(draws)
```

## Estimate

Fit/train model using observed data

### Grid approximation

```{r}
# randomly sample data from population
d.smp <- d.pop %>% slice_sample(n = sample.size)

H <- d.smp$height # observed covariates
W <- d.smp$weight # observed outcomes

# estimate using grid approximation
post <- ss %>%
  rowwise() %>%
  mutate(
    like = exp(sum(dnorm(W, mean = beta0 + beta1 * H, sigma, log = TRUE))),
    post.unnorm = like
  ) %>%
  ungroup() %>%
  mutate(
    post = post.unnorm / sum(post.unnorm)
    
  )

post
```

```{r}
draws <- post %>%
  slice_sample(n = n, weight_by = post, replace = TRUE) %>%
  select(beta0, beta1, sigma) %>% 
  posterior::as_draws_df()
```

```{r}
mcmc_hist(draws, pars = "beta1", binwidth = 6/50)
```

```{r}
summary(draws)
```

### Quadratic approximation

```{r}
m <- alist(
    W ~ dnorm(mu, sigma),
    mu <- beta0 + beta1 * H,
    beta0 ~ dnorm(0, 25),
    beta1 ~ dnorm(0, 1),
    sigma ~ dexp(0.3)
)
```

```{r}
f <- quap(m, data = d.smp)
```

```{r}
summary(f)
```

### Bayesian linear model estimation

```{r}
#| output: false
bayes.est.unif <- stan_glm(weight ~ 1 + height, data = d.smp)
```

```{r}
#| output: false
bayes.est.winf <- stan_glm(weight ~ 1 + height, data = d.smp,
                           prior = normal(0, 1),
                           prior_intercept = normal(0, 25),
                           prior_aux = exponential(0.3))
```

### Maximum likelihood linear model estimation

```{r}
ml.est <- lm(weight ~ 1 + height, data = d.smp)
```

### Compare models

```{r}
modelsummary::modelsummary(list(mle = ml.est, bayes.unif = bayes.est.unif, bayes.winf = bayes.est.winf), metrics = "all")
```

```{r}
estimate_slopes(bayes.est.unif, trend = "height")
```

```{r}
estimate_slopes(ml.est, trend = "height")
```


## Check model assumptions

XXX

## Test model using training data

XXX

## Test model using new data (replication)

XXX

## Profit but keep testing


## Population parameters

```{r}
f <- quap(m, data = d.pop)
```

```{r}
summary(f)
```

```{r}
#| output: false
bayes.est.unif <- stan_glm(weight ~ 1 + height, data = d.pop)
```

```{r}
#| output: false
bayes.est.winf <- stan_glm(weight ~ 1 + height, data = d.pop,
                           prior = normal(0, 1),
                           prior_intercept = normal(0, 25),
                           prior_aux = exponential(0.3))
```

```{r}
ml.est <- lm(weight ~ 1 + height, data = d.pop)
```

```{r}
modelsummary::modelsummary(list(mle = ml.est, bayes.unif = bayes.est.unif, bayes.winf = bayes.est.winf), metrics = "all")
```

```{r}
estimate_slopes(bayes.est.unif, trend = "height")
```

```{r}
estimate_slopes(ml.est, trend = "height")
```

