---
title: "rethinking_lm"
format:
  html:
    df-print: paged
---

## Setup environment

```{r}
#| output: false
library(tidyverse)
library(coursekata)
library(rethinking)

theme_set(theme_bw())
```

## Load data

```{r}
data(Howell1, package = "rethinking")
```


## Explore data

```{r}
Howell1
```


```{r}
gf_point(weight ~ height, data = Howell1, alpha = 0.5, color = 4) %>%
  gf_smooth(linewidth = 1, color = 2, method = "loess") %>% 
  gf_labs(x = "Height (cm)", y = "Weight (kg)")
```

```{r}
d.obs <- Howell1 %>% filter(age >= 18) %>% select(-c(male, age))

d.obs
```


```{r}
gf_point(weight ~ height, data = d.obs, alpha = 0.5, color = 4) %>%
  gf_smooth(linewidth = 1, color = 2, method = "loess") %>% 
  gf_labs(x = "Height (cm)", y = "Weight (kg)")
```

## Workflow

1.  State a clear question/goal
2.  Specify causal model (use a DAG to sketch your causal assumptions)
3.  Use causal model to build generative model
4.  Specify estimand (i.e., the unknown quantity of interest)
5.  Use generative model to build an estimator
6.  Test estimator using data simulated with generative model
7.  Apply estimator to observed data
8.  Check model assumptions
9.  Test model using observed data (posterior predictive check)
10. Test model using new data
11. Profit


## State a clear question/goal

-   What is the relationship between height and weight in adults?

-   Describe the association between height and weight in adults

## Specify causal model (DAG)

```{mermaid}
graph LR;
    H --> W;
```

$$\text{W} = f(\text{H})$$


## Use causal model to build generative model

The likelihood of the general linear model:

$$
y_i \stackrel{\text{i.i.d.}} \sim \text{Normal}(\mu = \beta_0 + \beta_1 \cdot x_i, \sigma)
$$

or (equivalently):

$$
y_i = \beta_0 + \beta_1 \cdot x_i + \epsilon_i
$$
$$
\epsilon_i \stackrel{\text{i.i.d.}} \sim \text{Normal}(0, \sigma)
$$

```{r}
sim_weight <- function(H, beta1, sigma) {
  U <- rnorm(length(H), mean = 0, sd = sigma)
  W <- beta1 * H + U
  return(W)
}
```

```{r}
H <- runif(500, 130, 180)
W <- sim_weight(H, 0.33, 5)
```

```{r}
d.sim <- tibble(height = H, weight = W)
```

```{r}
gf_point(weight ~ height, data = d.sim, alpha = 0.5, color = 3) %>%
  gf_smooth(linewidth = 1, color = 2, method = "loess") %>%
  gf_labs(x = "Height (cm)", y = "Weight (kg)")
```

```{r}
m <- alist(
    W ~ dnorm(mu, sigma),
    mu <- beta0 + beta1 * H,
    beta0 ~ dnorm(0, 10),
    beta1 ~ dnorm(0, 10),
    sigma ~ dunif(0, 10)
)
```

```{r}
f <- quap(m, data = d.sim)
```

```{r}
summary(f)
```


## Grid approximation

```{r}
beta0 <- seq(-10, 10, length.out = 50)
beta1 <- seq(-5, 5, length.out = 100)
sigma <- seq(0.1, 10, length.out = 50)

ss <- expand_grid(beta0, beta1, sigma)

H <- c(1, 2, 3)   # observed covariates
W <- c(5, 6, 7)   # observed outcomes

H <- c(130, 150, 170)   # observed covariates
W <- sim_weight(H, beta1 = 0.33, sigma = 5)   # observed outcomes

ss %>%
  rowwise() %>%
  mutate(
    likelihood.unnorm = exp(sum(dnorm(W, mean = beta0 + beta1 * H, sigma, log = TRUE)))
  ) %>%
  ungroup() %>%
  mutate(
    likelihood = likelihood.unnorm / sum(likelihood.unnorm)
  ) %>%
  pivot_longer(
    cols = c(beta0, beta1, sigma),
    names_to = "param",
    values_to = "value"
  ) %>% 
  gf_histogram(~ value, data = ss_long, weight = ~ likelihood) %>%
  gf_facet_wrap(~ param) %>%
  gf_labs(title = "Posterior distribution of parameters", x = "Parameter", y = "p")
```
