---
title: "Logistic regression/machine learning in R"
date: 2024-03-29
author: Edoardo \"Dado\" Marcora
execute:
  warning: false
format:
  html:
    toc: true
    df-print: kable
---

## Introduction

XXX

## Setup environment

```{r}
#| output: false
library(tidyverse)
library(coursekata)
library(performance)
library(parameters)
library(modelsummary)
library(ggeffects)
library(skimr)

theme_set(theme_bw())

set.seed(123)

options(scipen = 0, ggeffects_margin = "empirical")
```

## Background information

We will use the [Titanic dataset](https://www.kaggle.com/c/titanic/data) from [Kaggle](https://www.kaggle.com/) to predict survival on the Titanic and get familiar with logistic regression and machine learning.

### **Data dictionary**

+--------------+-------------------------------------------+---------------------------------------+
| **Variable** | **Definition**                            | **Key**                               |
+==============+===========================================+=======================================+
| `Survived`   | Survival                                  | `0` = No, `1` = Yes                   |
+--------------+-------------------------------------------+---------------------------------------+
| `Pclass`     | Ticket class                              | `1` = 1^st^, `2` = 2^nd^, `3` = 3^rd^ |
+--------------+-------------------------------------------+---------------------------------------+
| `Name`       | Name                                      |                                       |
+--------------+-------------------------------------------+---------------------------------------+
| `Sex`        | Sex                                       |                                       |
+--------------+-------------------------------------------+---------------------------------------+
| `Age`        | Age in years                              |                                       |
+--------------+-------------------------------------------+---------------------------------------+
| `SibSp`      | \# of siblings/spouses aboard the Titanic |                                       |
+--------------+-------------------------------------------+---------------------------------------+
| `Parch`      | \# of parents/children aboard the Titanic |                                       |
+--------------+-------------------------------------------+---------------------------------------+
| `Ticket`     | Ticket number                             |                                       |
+--------------+-------------------------------------------+---------------------------------------+
| `Fare`       | Passenger fare                            |                                       |
+--------------+-------------------------------------------+---------------------------------------+
| `Cabin`      | Cabin number                              |                                       |
+--------------+-------------------------------------------+---------------------------------------+
| `Embarked`   | Port of embarkation                       | `C` = Cherbourg\                      |
|              |                                           | `Q` = Queenstown\                     |
|              |                                           | `S` = Southampton                     |
+--------------+-------------------------------------------+---------------------------------------+

### **Variable notes**

**Pclass**: A proxy for socio-economic status (SES)\
1^st^ = Upper\
2^nd^ = Middle\
3^rd^ = Lower\
\
**Age**: Age is fractional if less than 1. If the age is estimated, is it in the form of xx.5\
\
**SibSp**: The dataset defines family relations in this way...\
Sibling = brother, sister, stepbrother, stepsister\
Spouse = husband, wife (mistresses and fiancés were ignored)\
\
**Parch**: The dataset defines family relations in this way...\
Parent = mother, father\
Child = daughter, son, stepdaughter, stepson\
Some children travelled only with a nanny, therefore parch=0 for them

## Read and prep data

```{r}
d1 <- read_csv("train.csv", show_col_types = FALSE)

head(d1)
```

```{r}
skim(d1)
```

```{r}
d1 <- d1 %>%
  select(-c(PassengerId, Name, Ticket, Cabin)) %>%
  drop_na(Embarked) %>%
  mutate(
    Survived = factor(Survived),
    Age = if_else(is.na(Age), median(Age, na.rm = TRUE), Age),
    Sex = factor(Sex),
    Pclass = factor(
      Pclass,
      levels = c(1, 2, 3),
      labels = c("1st", "2nd", "3rd"),
      ordered = TRUE),
    Embarked = factor(
      Embarked,
      levels = c("C", "Q", "S"),
      labels = c("Cherbourg", "Queenstown", "Southampton"),
      ordered = FALSE))
```

```{r}
skim(d1)
```

## Exploratory data analysis

```{r}
gf_props(~ Survived, fill =  ~ Sex, position = "fill", data = d1)
```

```{r}
gf_props(~ Survived, fill =  ~ Pclass, position = "fill", data = d1)
```

```{r}
gf_props(~ Survived, fill =  ~ Embarked, position = "fill", data = d1)
```

```{r}
gf_density(~ Age, fill = ~ Survived, data = d1)
```

```{r}
favstats(Age ~ Survived, data = d1)
```

```{r}
gf_density(~ log(Fare + 1), fill = ~ Survived, data = d1)
```

```{r}
favstats(Fare ~ Survived, data = d1)
```

## Fit simple linear regression model to the data

```{r}
m1 <- glm(Survived ~ Age + Sex + Pclass, family = "binomial", data = d1)
```

```{r}
#| fig-height: 8
check_model(m1)
```

## Evaluate model and compute effect sizes

```{r}
modelsummary(m1, stars = TRUE)
```

```{r}
model_parameters(m1, exponentiate = TRUE)
```

```{r}
pred.sex <- predict_response(m1, "Sex")

pred.sex
```

```{r}
test_predictions(pred.sex)
```

```{r}
pred.pclass <- predict_response(m1, "Pclass")

pred.pclass
```

```{r}
test_predictions(pred.pclass)
```

```{r}
pred.age <- predict_response(m1, "Age")

pred.age
```

```{r}
test_predictions(pred.age)
```

## Print environment

```{r}
sessioninfo::session_info()
```
